---
title: "STA5092Z: Assigment 1"
author: "Yevashan Perumal"
date: "3/30/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(gridExtra)
```

## Summary of data
Data were collected for each austral summer breeding season from September 2005February 2019 (14 breeding seasons in total).
The data consists of 6 csv files; 3 files containing data and 3 files containing metadata describing the fields in each data set. Below we import, wrangle and explore some of the data.

## Part 1 - Data Wrangling
```{r Import Data q2, include=FALSE}
# First Import Nestling comes first
df1 <- read.csv("./Datasets/Data.Bourne_NestlingsAllData.csv", header = TRUE)
df1 <- tibble(df1) #Convert to tibble

# Second import Fledgeling comes second
df2 <- read.csv("./Datasets/Data.Bourne_FledglingSurvival.csv", header = TRUE)
df2 <- tibble(df2) #Convert to tibble

# Third Import
df3 <- read.csv("./Datasets/Data.Bourne_NestSuccess.csv", header = TRUE)
df3 <- tibble(df3)  #Convert to tibble
```

A preview of the dataset named Data.Bourne_NestlingsAllData.csv
```{r df1, echo=FALSE}
head(df1,5)
# sum(is.na(df1$Date)) #No nulls in the date column
```
\newpage
A preview of the dataset named Data.Bourne_FledglingSurvival.csv
```{r df2 ,echo=FALSE} 
head(df2,5)
# sum(is.na(df2$Date)) #No nulls in the date column
```
A preview of the dataset named Data.Bourne_NestSuccess.csv:
```{r df3 ,echo=FALSE}
head(df3,5)
```
Before merging the Nestling and Fledgling dataframes, several actions need to be taken to cleanse the data and make it ready. They are detailed below:

### Fledgeling data
* Convert all date related fields from character  to date format
* Remove any potential whitespace from character fields
* convert Surv1Mass,GrpSizeAD,GrpSizeTotal and X.Imm to numeric type from character.
* Renaming GRP to Group

\newpage
### Nestling data
* Convert date related field to date format
* Convert Surv1Mass to numeric
* Remove whitespaces from character fields

It does appear from NAs were introduced during the data conversion of characater to numeric, which are for fields that did not have any data to start with.

```{r clean up dataframes q1.3, include=FALSE,echo=FALSE}
df1 <- df1 %>% 
    rename(Group=GRP)%>% 
    mutate(Date = dmy(Date),
           TwoMonthsPrior =dmy(TwoMonthsPrior),
           IncDate =dmy(IncDate),
           HatchDate =dmy(HatchDate),
           FledgeDate =dmy(FledgeDate),
           FailDate =dmy(FailDate),
           IndepDate =dmy(IndepDate),
           X1YrDate =dmy(X1YrDate),
           X2YrDate =dmy(X2YrDate),
           Surv1Mass=as.numeric(Surv1Mass),
           BirdID = str_replace_all(BirdID, " ", ""),
           Group=str_replace_all(Group, " ", ""),
           Season=str_replace_all(Season, " ", ""),
           NestID=str_replace_all(NestID, " ", ""),
           GrpSizeAD=as.numeric(GrpSizeAD),
           GrpSizeTotal=as.numeric(GrpSizeTotal),
           X.Imm=as.numeric(X.Imm)
           )%>%
    arrange(BirdID,Date,Group,Season)

df2 <- df2 %>%
    mutate(Date = dmy(Date),
           Surv1Mass=as.numeric(Surv1Mass),
           BirdID = str_replace_all(BirdID, " ", ""),
           Group=str_replace_all(Group, " ", ""),
           Season=str_replace_all(Season, " ", ""),
           NestID=str_replace_all(NestID, " ", ""),) %>% 
    arrange(BirdID,Date,Group,Season)
```

```{r merge q1.3, include=FALSE}
#V1
# Spiedbabbler <- df1%>% left_join(df2,by=c('Date','BirdID','Group','Season','NestID'))

# V2
Spiedbabbler <- df1%>% left_join(df2)
# head(Spiedbabbler,5)
# print(dim(df1))
# print(dim(df2))
# print(dim(Spiedbabbler))
#Nestling comes first!
d <- dim(Spiedbabbler)
```
The dimensions of the merged data set are  `r d[1]` rows and `r d[2]` columns.
The nestling data was used as the base as this is the earlier developmental stage before they become fledglings.Some birds do not make it to fledgling thus the nestling data has more rows/birds. Therefore, the fledgling data is left-joined to nestlings to get the most complete picture possible.

## Part 2
2.1)The Total Sample Size Per Year:
```{r q2.1, echo=FALSE}
#Total Sample Size Per Year
size_per_year <-Spiedbabbler %>%
    mutate(year=year(Date)) %>%
    select(c(year,BirdID))%>%
    group_by(year)%>%
    summarise('Sample Size' = n())
    # summarise(sample_size==count(BirdID))


# print(size_per_year)
#Print as a table
knitr::kable(
  size_per_year, 
  caption = "The Total Sample Size Per Year"
)
```
2.2.1) Checking for Nulls
```{r q2.2 Nulls,echo=FALSE}
#Check for nulls per column
miss_val <-Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    sapply(function(x) sum(is.na(x)))

#Print as table
knitr::kable(
  miss_val, 
  caption = "Number of Missing Values in Selected Fields",
  col.names =c("Count")
)

```
The occurance of nulls indicates we need to be cater for them when plotting or modeling with these
fields.
\newpage
2.2.2)Boxplots to allow us to look for outliers:

```{r q2.2 outliers, echo=FALSE}
# Create a boxplot for each of the variable of interest
b1<-Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=GrpSizeAD))+
    geom_boxplot(na.rm = T)

b2 <- Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=GrpSizeTotal))+
    geom_boxplot(na.rm = T)

b3<-Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=BroodSize))+
    geom_boxplot(na.rm = T)

b4 <- Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=Mass))+
    geom_boxplot(na.rm = T)

b5<-Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=TMaxMeas))+
    geom_boxplot(na.rm = T)

b6 <- Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=No.ChicksFledge))+
    geom_boxplot(na.rm = T)

b7<-Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=TotRain))+
    geom_boxplot(na.rm = T)

b8 <- Spiedbabbler %>%
    select(c(GrpSizeAD,GrpSizeTotal,BroodSize,Mass,TMaxMeas,
             No.ChicksFledge,TotRain,Drought))%>%
    ggplot(aes(y=Drought))+
    geom_boxplot(na.rm = T)

#Place all boxplots on to one plot
grid.arrange(b1,b2,b3,b4,b5,b6,b7,b8, nrow = 2,ncol=4,top = "Boxplots of Variables to Find Outliers")
```

An outlier is an observation that lies at or further away than 1.5 times the interquartile range, either below the lower quartile or above the upper quartile i.e. any observations further out than the "whiskers' on a box plot.

The following variables appear to have 1 or more outliers:

* GrpSizeAD
* GrpSizeTotal
* BroodSize
* Mass
* TMaxMeas

The following variable appear to not have outliers:

* No.ChicksFledge
* TotRain

The variable Drought appears to be a binary indicator (1/0) as to whether a drought occurred, and thus would not have any outliers; despite what the boxplot appears to indicate "1" is a valid observation and not an outlier

2.3)
```{r q2.3 per nest per date mass, echo=FALSE, message=FALSE}
#Code to find total mass per nest per date
mass_by_nest<-Spiedbabbler%>%
    select(NestID,Date,Mass)%>%
    group_by(NestID,Date)%>%
    summarise('Total Mass'=sum(Mass))%>%
    arrange(NestID,Date)
#Print result in a table
knitr::kable(
  head(mass_by_nest,5), 
  caption = "Preview of Mass and Date per Nest"
)
```
The dimensions of the table are `r dim(mass_by_nest)[1]` rows and `r dim(mass_by_nest)[2]` columns.

2.4)
```{r q2.4 ,echo=FALSE} 
# Spiedbabbler%>%
#     select(Season,TMaxMeas,TMinMeas,TvarMeas)
#     na.omit()%>%
   # summarise(average_high_temp=mean(TMaxMeas),average_low_temp=mean(TMinMeas))%>%
# sum(is.na(Spiedbabbler$TvarMeas))

diff <-Spiedbabbler%>%
    select(Season,TvarMeas,TMaxMeas,TMinMeas)%>%
    na.omit()%>% #There are null values for some observations
    group_by(Season)%>%
    summarise('Avg Tmp Range'=mean(TvarMeas),'
              Max Temp'=max(TMaxMeas),
              'Min Temp'=min(TMinMeas))

diff


# %>%
#     mutate(rn=row_number())
# knitr::kable(
#   diff,
#   caption = "Comparison of Temperature Measures per Season",
#   digits=2
# )


# plto_labs <- list(diff$Season)
# typeof(plto_labs)
# diff %>% ggplot(aes(x=factor(rn),y=mean_change))+
#           geom_point(aes(type='l')) +
#           scale_x_discrete(breaks = 1:13, labels=plto_labs)
```
From the table we see that both maximum and minimum temperatures appear to be on a downward trend. However the average daily temperature variability does not appear to any significant trends that we can pick out.

2.5)

```{r q2.5, echo=FALSE, message=FALSE}
#Create the required plot
Spiedbabbler%>%
    select(NestID,Date,Mass,Season)%>%
    group_by(NestID,Date,Season)%>%
    summarise(sum_mass=sum(Mass))%>%
    ggplot(aes(x=Date,y=sum_mass,color=Season))+
    geom_point(na.rm = T)+
    ylab("Season")+
    labs(title = "Average Mass per Nest Over Time,faceted per Season")+
    facet_wrap(~Season,scales='free_x')

```
*Apologies for the squashed axis on the graphics.
\newpage
2.6)

```{r q2.6, echo=FALSE, message=FALSE}
#Create year column,find averages, then craete a facet on Season for each variable.

p1 <-Spiedbabbler%>%
    mutate(year=year(Date))%>%
    select(year,Mass,BroodSize,GrpSizeTotal,No.ChicksFledge,TotRain,Season,Drought)%>%
    na.omit%>%
    group_by(year,Season)%>%
    summarise(avg_mass=mean(Mass),mean_broodsize=mean(BroodSize),
              avg_grp_size=mean(GrpSizeTotal),avg_fledg_count=mean(No.ChicksFledge),
              avg_rain=mean(TotRain))%>%
    ggplot(aes(x=year,y=avg_mass))+
    geom_bar(stat="identity")+
    facet_wrap(~Season,scales = 'free_y')+
    xlab("Year")+
    ylab("Average Mass")

p2 <-Spiedbabbler%>%
    mutate(year=year(Date))%>%
    select(year,Mass,BroodSize,GrpSizeTotal,No.ChicksFledge,TotRain,Season,Drought)%>%
    na.omit%>%
    group_by(year,Season)%>%
    summarise(avg_mass=mean(Mass),mean_broodsize=mean(BroodSize),
              avg_grp_size=mean(GrpSizeTotal),avg_fledg_count=mean(No.ChicksFledge),
              avg_rain=mean(TotRain))%>%
    ggplot(aes(x=year,y=mean_broodsize))+
    geom_bar(stat="identity")+
    facet_wrap(~Season,scales = 'free_y')+
    xlab("Year")+
    ylab("Average Brood Size")
              

p3 <-Spiedbabbler%>%
    mutate(year=year(Date))%>%
    select(year,Mass,BroodSize,GrpSizeTotal,No.ChicksFledge,TotRain,Season,Drought)%>%
    na.omit%>%
    group_by(year,Season)%>%
    summarise(avg_mass=mean(Mass),mean_broodsize=mean(BroodSize),
              avg_grp_size=mean(GrpSizeTotal),avg_fledg_count=mean(No.ChicksFledge),
              avg_rain=mean(TotRain))%>%
    ggplot(aes(x=year,y=avg_grp_size))+
    geom_bar(stat = "identity")+
    facet_wrap(~Season,scales = 'free_y')+
    xlab("Year")+
    ylab("Average Total Group Size")
                
p4 <-Spiedbabbler%>%
    mutate(year=year(Date))%>%
    select(year,Mass,BroodSize,GrpSizeTotal,No.ChicksFledge,TotRain,Season,Drought)%>%
    na.omit%>%
    group_by(year,Season)%>%
    summarise(avg_mass=mean(Mass),mean_broodsize=mean(BroodSize),
              avg_grp_size=mean(GrpSizeTotal),avg_fledg_count=mean(No.ChicksFledge),
              avg_rain=mean(TotRain))%>%
    ggplot(aes(x=year,y=avg_fledg_count))+
    geom_bar(stat = "identity")+
    facet_wrap(~Season,scales = 'free_y')+
    xlab("Year")+
    ylab("Average No. of Fledgling Chicks")

p5 <-Spiedbabbler%>%
    mutate(year=year(Date))%>%
    select(year,Mass,BroodSize,GrpSizeTotal,No.ChicksFledge,TotRain,Season,Drought)%>%
    na.omit%>%
    group_by(year,Season)%>%
    summarise(avg_mass=mean(Mass),mean_broodsize=mean(BroodSize),
              avg_grp_size=mean(GrpSizeTotal),avg_fledg_count=mean(No.ChicksFledge),
              avg_rain=mean(TotRain))%>%
    ggplot(aes(x=year,y=avg_rain))+
    geom_bar(stat = "identity")+
    facet_wrap(~Season,scales = 'free_y')+
    xlab("Year")+
    ylab("Average Total Rainfall")

# str(Spiedbabbler)
```

```{r echo=FALSE ,fig.height=4, fig.width=10}
p1+labs(title="Average Mass per Year by Season")
```
```{r echo=FALSE,fig.height=4, fig.width=10}
p2+labs(title="Average Brood Size per Year by Season")
```
```{r echo=FALSE,fig.height=4, fig.width=10}
p3+labs(title="Average Total Group Size per Year by Season")
```
```{r echo=FALSE,fig.height=4, fig.width=10}
p4+labs(title="Average No. of Fledgling Chicks per Year by Season")
```
```{r echo=FALSE,fig.height=4, fig.width=10}
p5+labs(title="Average Total Rainfall per Year by Season")
```
2.7) The study this data was collected for was investigating if the survivability of the young birds at different developmental stages is affected by environmental conditions and/or group size.
Therefore it makes sense to plot a graphic with an environmental factor such as temperature against the number of young that survive.

```{r q2.7, echo=FALSE}
# Spiedbabbler%>%
#   mutate(year=year(Date))%>%
#   select(year,BirdID,Sex,TMaxMeas)%>%
#   group_by(year,Sex)%>%
#   ggplot(aes(x=year,fill=Sex))+
#   geom_bar(position = "dodge")+
#   labs(title = "Distrubtion of Observed Sex over Time")+
#   ylab('Number of Birds')+
#   xlab("Years")+
#   geom_line(aes(y=TMaxMeas))

sum(is.na(Spiedbabbler$SurvInd))
unique(Spiedbabbler$SurvInd)

Spiedbabbler%>%
  select(Season,SurvInd,TMaxMeas)%>%
  na.omit%>% #Remove NA observations
  group_by(Season)%>%
  summarise(sum_surv=sum(SurvInd),mean_max_temp=mean(TMaxMeas))%>%
  ggplot(aes(x=Season))+
  geom_bar(aes(y=sum_surv),stat = "identity",fill="red",alpha=0.4)+
  geom_point(aes(y=mean_max_temp,col=mean_max_temp))+
  ylab('No of Chicks Surviving to 90days')+
  labs(title = "Number of Chicks Surviving to Nutritional Independance by Season")+
  scale_color_continuous(name="Max Temp")


```
From the figure above, it does appear that higher maximum temperatures could affect the survivability of young birds. The two seasons with the highest survivability numbers are those with the lowest maximum temperature. However, this is not concrete enough evidence and warrants further investigation and testing.


